---
date: '2012-12-03'
title: On the fly bam to sam conversion using named pipes
tags: [R]
comments: true
---

In bioinformatics the common format developed for storing short read alignments
is the [SAM](http://samtools.sourceforge.net/SAM1.pdf) format which has a binary
representation and an ASCII text form.  There exists a C API to work with the
binary format directly, as well as language bindings for most of the common
programming languages.  Heng Li, the author of the format and the bwa
short read aligner, created the samtools program to work with the SAM format and
convert between bam/sam among many other tasks.  However often third-party
programs will only read the ASCII SAM files, which typically have a .sam
extension, rather than the binary files, with a .bam extension.  In addition,
the sam files are completely uncompressed, so can take upwards of 3x or more
disk space than the compressed BAM files.  This gives us motivation to avoid
having uncompressed SAM files at any point, even as a temporary file which will
be deleted.

If the third-party program in question has the ability to accept input from
standard input, the solution is very straightforward.

```{r eval = F, engine = 'bash'}
samtools view file.bam | my_program --arguments
```

However, if the program can only accept named files, often people think the only
option is to create the temporary file. Luckily, this is not the case, and
linux has long had functionality to treat a pipe as though it were a file. So
there exists a very clean solution to the problem.

```{r eval = F, engine = 'bash'}
mkfifo file.sam; samtools view file.bam > file.sam& my_program --arguments file.sam; rm file.sam
```

This creates a named pipe using mkfifo, converts the file using samtools view
and puts that in the background, the runs the third-party program with the named
pipe, then removes the pipe once the program is complete.  The sam file is never
stored on disk.
