---
date: '2015-05-20'
title: covr - Test Coverage for R Packages
tags: [R, covr]
projects: [covr]
comments: true
---

Being primarily a statistical language R lacks a number of common code analysis
tools available for languages used more often for general programming.  Testing
has been done in the R source from [very early
on](https://github.com/wch/r-source/tree/7685a96319d94bac3d82e06eff4a469feec72a9a/tests)
(Martin Maechler started adding tests on April 9th, 1998, 17 years ago!)
However most of this testing was ad hoc, either requiring visual examination of
the outputs, or simply testing current vs previous outputs.  The
[RUnit](http://cran.r-project.org/web/packages/RUnit/index.html) (2005),
[svUnit](http://cran.r-project.org/web/packages/svUnit/index.html) (2009) and
[testthat](http://cran.r-project.org/web/packages/testthat/index.html) (2009)
packages brought more formal unit testing into R community.

So while tools existed to show authors _how_ to test, there were very limited
tools to show _what_ to test.  Code coverage has a very long history ([Miller
et al. 1963](http://dl.acm.org/citation.cfm?doid=366246.366248)) in the
software development community and remains a useful metric today. Unfortunately
the R community has had very few options to track and report on code coverage.
[R-coverage](http://r2d2.quartzbio.com/posts/r-coverage-docker.html) by Karl
Forner and
[testCoverage](https://github.com/MangoTheCat/testCoverage) by Tom Taverner,
Chris Campbell, Suchen Jin were two existing solutions, however both were
somewhat problematic to use.  The former required a patch to the R source,
which is a non-starter for widespread use (particularly on Windows).  The
latter had a complicated implementation which was difficult to use
interactively and supported only a limited output format.  Neither project
supported tracking coverage of compiled code in addition to R source or
supported coverage tracking services such as <http://codecov.io> or
<http://coveralls.io>.

Covr aims to make code coverage both easy to use and informative in R.  Package
coverage can be tracked and tests run very simply

```{r eval = F}
library(covr)
cov <- package_coverage()
```

The results can be summarized with an informative `print()` method, as well as
coerced to a `data.frame`.  There is also an interactive Shiny App available
via `shine()` which can be used to inspect the code visually and identify
coverage gaps.

In addition covr supports both the <http://codecov.io> and
<http://coveralls.io> coverage reporting services.  These services will track
the coverage of a project over time and report on coverage changes as well as
exploration of the source code.  They are designed to work seamlessly within a
CI system such as <https://travis-ci.org/> or <http://wercker.com/>.  Simple
functions exist to instrument the code, run tests and submit the results for
both services and can easily be added to the CI build script to automatically
run coverage for every build.

```{r, eval = FALSE}
library(covr)
# for codecov.io
codecov()

# for coveralls.io
coveralls()
```

Covr version 1.0.0 is now [on CRAN](http://cran.r-project.org/web/packages/covr/index.html)
and Covr development is ongoing at <https://github.com/jimhester/covr>, pull
requests for additional features or bug fixes are always welcome!
