---
date: '2014-02-12'
title: ZSH global aliases
tags: [R]
comments: true
---

Just a short post about two useful global aliases I created.  ZSH global
aliases are basically variables which are expanded before the command is
executed.  This allows them to be placed anywhere on the line, not just at the
start like traditional aliases.  Bash (as far as I know) does not have an
analog to ZSH global aliases, but I have found them very useful.

Both of them produce exactly the same output (unique lines in a file), but in two different ways.

```{r eval = F, engine = 'bash'}
alias -g U="awk '!a[\$0]++'"
alias -g SU='\(sort | uniq\)'
```

The first uses awk to hash the lines seen, and only print the line if the current
line is not in the hash.  This finds the unique lines without sorting, which
preserves the original order is usually much much faster.  The only issue is it
can exhaust system memory if used on extremely large files.

How much faster is hashing for uniques rather than sorting for uniques?

*Note these commands use GNU shuf to randomly shuffle*

```{r eval = F, engine = 'bash'}
gshuf /usr/share/dict/words | head -n 100000 > words #get 100000 random words
(for i in {1..100};do gshuf words | head -n 1000;done;cat words) | gshuf > dup_words #randomly pick 1000 words to duplicate 100 times, append the full list and randomize
```

Now lets count unique lines in the files

```{r eval = F, engine = 'bash'}
time awk '!a[$0]++'< dup_words | wc -l
#  100000
#  awk '!a[$0]++' < dup_words  0.14s user 0.00s system 99% cpu 0.149 total
#  wc -l  0.00s user 0.00s system 2% cpu 0.147 total
time (sort | uniq) < dup_words | wc -l
#  100000
#  ( sort | uniq; ) < dup_words  2.45s user 0.01s system 100% cpu 2.438 total
#  wc -l  0.00s user 0.00s system 0% cpu 2.437 total
```

So you get the same results 16X faster using the hashing strategy.

And lets use the aliases which started this whole mess

```{r eval = F, engine = 'bash'}
U < dup_words | wc -l
SU < dup_words | wc -l

cat dup_words | U | wc -l
cat dup_words | SU | wc -l
```

All in all this allows you to do sorting faster and with less typing than the
ubiquitous `| sort | uniq` pattern.
