---
date: '2012-11-08'
title: External Git Repositories
tags: [R]
comments: true
---

It is sometimes useful to have git tracking files stored in a different 
location from your working directory, for instance if you would like to backup 
the git files, and your working directory contains files which are much too 
large to back up.  This could be accomplished by using symlinks, however this 
is a fairly fragile solution.  A more robust solution is that given in [this 
stackoverflow post][so].

If you already have a git repository you would like to store outside the 
working directory, say /path/to/repo.git and you are in the working directory

```{r eval = F, engine = 'bash'}
mv .git /path/to/repo.git
echo "gitdir: /path/to/repo.git" > .git
git config --add worktree `pwd -P`
```

This will move the .git folder to the new location, create a .git file pointing 
to the new location, and update the git configuration to point to the current 
working directory.  Everything will then work exactly as it did before

If you are initializing a new empty repository you can do everything on one 
line

```{r eval = F, engine = 'bash'}
git --git-dir=/path/to/repo.git --work.tree=`pwd -P` init && \
  echo "gitdir: /path/to/repo.git" > .git
```

This is a very useful tip for keeping track of all of your repositories, and 
also allows easy backup, as you can just mirror the folder which contains all 
of them.

[so]: http://stackoverflow.com/questions/505467/can-i-store-the-git-folder-outside-the-files-i-want-tracked
